---
title: "Interpreting Social Patterns with Census Data"
author: "John Lauermann, School of Information, Pratt Institute"
date: "Last updated: November 2025"
output: github_document
---

# Overview

Census data products are the primary historic record of American society, available as open data for every community in the United States. The long temporal record and diverse geographic scope of these data present opportunities for data storytelling at scale. Census data allow you to tell stories that move beyond small scale case studies, to think about national patterns and historic trends.

In this workshop, you'll learn how to visualize and analyze Census data in ways that could support data storytelling. The script that follows proceeds in three parts:

1.  Explore and query Census data from the Census API using `tidycensus`

2.  Visualize data patterns using `ggplot2` and a data visualization approach called the "[grammar of graphics](https://www.tandfonline.com/doi/abs/10.1198/jcgs.2009.07098)"

3.  Statistically test data patterns using R tools for correlation and simple linear regression.

#### Data storytelling context

In the workflow that follows, I'll the topic of housing affordability. My primary variable of interest is median rent – a key metric for assessing the scope of the housing crisis in American cities. There are two broad schools of thought around why housing is so unaffordable. In this example I'd like to do some data storytelling around these potential story lines.

-   One approach focuses on a lack of [housing supply](https://www.urban.org/projects/road-map-address-americas-housing-crisis). The supply-side approach is common in the [YIMBY movement](https://onlinelibrary.wiley.com/doi/abs/10.1111/1468-2427.13062) and the '[abundance agenda](https://www.simonandschuster.com/books/Abundance/Ezra-Klein/9781668023488)', political movements that argue for deregulating housing development through reforms to zoning and building codes. I'll measure that by assessing new home construction.

-   Another approach focuses on [housing demand](https://journals.sagepub.com/doi/abs/10.1177/0042098020940596). The demand-side argument is common among academic researchers and activist groups that highlight the [uniquely sticky](https://www.nber.org/papers/w22816) nature of housing markets, and [market failures](https://www.jchs.harvard.edu/press-releases/new-report-highlights-unease-housing-market-amid-worsening-affordability-crisis) like a tendency for developers to focus on luxury real estate when most demand is actually for affordable housing. I'll measure this by assessing newcomers within neighborhoods.

These factors are not mutually exclusive: the truth is likely that we need policy solutions focused on **both** supply and demand factors. The goal below is to present a more nuanced picture of the debate, by assessing patterns at scale and using statistical tests to think beyond simply visualizing patterns.

#### Getting started

Download this file from GitHub and then open it in R Studio (`File>Open File`). Choose the Visual view if it doesn't automatically populate.

To run the code, either click the small green "Run Current Chunk" button in the top right of each code chunk, or put your cursor anywhere in the chunk then hit `Cntrl + Shift + Enter` (for a PC) or `Cmd + Shift + Enter` (for a Mac).

We'll start by loading the necessary packages for this workflow. In R, a package is a collection of tools that are used for specific kinds of workflows. The packages are all open source, and you can synthesize as needed within your own code.

```{r}
# verify whether you need these packages, and install if so
if (!require(dplyr)) install.packages("dplyr", dependencies = TRUE)
if (!require(ggplot2)) install.packages("ggplot2", dependencies = TRUE)
if (!require(stringr)) install.packages("stringr", dependencies = TRUE)
if (!require(tidycensus)) install.packages("tidycensus", dependencies = TRUE)
if (!require(tmap)) install.packages("tmap", dependencies = TRUE)
```

```{r}
# now load the libraries
library(dplyr)      # for managing data tables
library(ggplot2)    # for data visualization
library(stringr)    # for filtering string columns
library(tidycensus) # for using the Census API
```

# Task 1: Explore and query Census data

The US Census Bureau publishes much of its data via the [Census Data API](https://www.census.gov/data/developers/guidance/api-user-guide.html). If you already know what you're looking for, the API is often a more efficient way to query data than by navigating and downloading from the general [Census website](https://data.census.gov/). That said, you *really* have to understand the structure of Census data in order to use the API. So in this section we'll briefly recap Census products, geographies, and structure of content. Potentially useful ways to explore Census data include:

-   The Census data website, which has an interactive query function: <https://data.census.gov/>

-   The Census API Discovery Tool, which explains the API structure: <https://api.census.gov/data.html>

We'll start by setting up an API call. To get started, you need to obtain a free [Census API key](https://api.census.gov/data/key_signup.html). Save this somewhere secure, and remember not to share it on public platforms (e.g. when you share your code on a portfolio...).

```{r setup, echo=FALSE}
# set your API key
census_api_key("your key here")
```

#### **Data products**

Most of the time, we are interested in the demographic, socioeconomic, housing, and employment data from the Decennial Census and American Community Survey.

The [Decennial Census](https://www.census.gov/programs-surveys/decennial-census.html) has been conducted every decade since 1790. It is the longest and most comprehensive record of American society, covering thousands of variables. Like any *census*, it is intended to be a comprehensive count of every person living in the country (though there are well-known issues around enumerating some ["hard-to-count populations"](https://www.census.gov/newsroom/blogs/random-samplings/2023/10/understanding-undercounted-populations.html)).

The [American Community Survey (ACS)](https://www.census.gov/programs-surveys/acs.html) is a companion to the Decennial Census, run every year since 2005. It includes many of the same variables counted in the Census, available on a more regularly updated basis. But like any *survey*, the ACS counts only a small sample of the overall population (typically about 3.5 million households annually, out of \~127 million households nationally). For this reason ACS data are always published in two forms, an *estimate* of the data value and a *margin of error* describing uncertainty about that data value. Furthermore, ACS data are aggregated into different temporal estimates:

-   *ACS 1 year estimates* use data from only the year in question. They are only available for some variables and at course geographic scales.

-   *ACS 3 year estimates* use data from the year in question plus the preceding two years (e.g. a 2021-2023 ACS 3 year estimate is an estimate of 2023 values, based on an average of data from 2021 through 2023). They are available for most variables and some finer geographic scales.

-   *ACS 5 year estimates* use data from the year in question plus the preceding four years (e.g. a 2019-2023 ACS 3 year estimate is an estimate of 2023 values, based on an average of data from 2019 through 2023). They are available for nearly all variables at nearly all geographic scales.

For this exercise, we'll work with data from the ACS 5 year estimates, for the most recent year available. To do that, we'll first pull up a list of all the variables in that data product. Here is a [list of options](https://walker-data.com/tidycensus/articles/basic-usage.html#searching-for-variables) for defining the parameters on that `load_variables()` function.

```{r}
# query metadata from the api, naming the year and data product
# parameter options available at https://walker-data.com/tidycensus/articles/basic-usage.html#searching-for-variables 
variablelist <- load_variables(year= 2023, dataset = "acs5")
```

You now have a data frame called `variablelist` under the Environment tab on the top right of your screen. You can click on it to open and see the structure of what we downloaded. This will list each census variable code, a variable name, and information on the sampling structure and finest scale at which data are available.

#### Querying the API

Once you know the identifier codes for the variable(s) you want, we can configure an API call. Remember that Census/ACS data come packaged in numerous geographic scales, each identified with nesting FIPS codes.

-   *States*: There are 50 of them, covering the entire US.

-   *Counties*: There are 3007 of them, nested within states.

-   *Tracts*: Each tract is a statistical area home to somewhere between 2500-8000 residents. They describe what the census considers to be a relatively meaningful population area based on demographic, economic, and housing characteristics. Their geographic size is inversely proportional to their population density, so they will be small in cities and larger in rural areas. They are nested within counties, which are nested within states.

-   *Block groups*: Each block group is a subdivision of a tract, home to between 600-3000 people. This is generally the finest scale of aggregation available for ACS data due to survey limitations and data privacy concerns.

Here is a list of [geography options](https://walker-data.com/tidycensus/articles/basic-usage.html#geography-in-tidycensus) for the `tidycensus` library. See also the Census [*Geographic Areas Reference Manual*](https://www2.census.gov/geo/pdfs/reference/GARM/) for all the technical details.

```{r}
# define a list of variables, including more intuitive names
variables <- c(MedianConRent = "B25058_001",
               HouseUnits_sum = "B25002_001",
               HouseUnits_built2020orlater = "B25034_002",
               OccupiedHouseUnits_sum = "B25038_001",
               Owner_MovedIn_after2021 = "B25038_003",
               Renter_MovedIn_after2021 = "B25038_010"
               )
```

```{r}
# set up an API query
data <- get_acs(geography = "county",    # all counties in the US
                variables = variables,   # all variables defined in the list above
                output = "wide",         # row = observation, column = variable
                year = 2023,             # most recent year
                geometry = TRUE         # include spatial boundary data   
                )             
```

You now have a data frame called `data` in your Environment. You can open it to see the structure of what we downloaded. Note that ACS data publishes both estimated values (variables will include 'E' at the end of the column name) and margins of error on those estimates (will include 'M' at the end of the column name).

```{r}
# see the total rows and columns of the data
dim(data)

#list the column names
ls(data)
```

Finally, we need to do a bit of data processing to get the data into a shape that will allow further visualization and analysis.

```{r}
# use dplyr tools to clean the data
data <- data %>%
  rename_with(~ gsub("E$", "", .x), .cols = everything()) %>%
  mutate(
    NewHomes_pct =  (HouseUnits_built2020orlater / HouseUnits_sum) * 100,
    NewComers_pct = (Owner_MovedIn_after2021+ Renter_MovedIn_after2021) /
      OccupiedHouseUnits_sum * 100)
```

```{r}
# now we'll calculate some descriptive statistics just to verify it worked
mean(na.omit(data$MedianConRent))
mean(na.omit(data$NewHomes_pct))
mean(na.omit(data$NewComers_pct))
```

# Part 2: Visualize data patterns

To visualize the data, we'll use a library called `ggplot2`. It is based on a statistical visualization concept called "[grammar of graphics](https://www.tandfonline.com/doi/abs/10.1198/jcgs.2009.07098)" (hence the 'gg' applied to a 'plot'). The grammar of graphics is a layering structure, in which you first define a plot space and then add on additional layers of information.

You build graphics by adding these layers, with a + for each additional layer:

-   **ggplot**() defines the plot layer itself, based on the **data** you want to use

-   **aes** defines the aesthetic mapping of a layer, connecting variables to visual elements of the plot

-   **geom** adds geometric layers onto the plot, from a library of potential data viz types

-   **stat** functions calculate derivative numbers for use in a graphic

-   **position** adjustments change how layers are ordered or rendered

-   **coordinate** functions alter how the points are displayed (e.g. changing the projection of a map)

-   **facet** functions allow you to arrange multiple charts together

-   **theme** borrows from a [library of themes](https://ggplot2.tidyverse.org/reference/ggtheme.html) to style the graphic

#### Histogram

We'll start by exploring the overall distribution of the rent data with a histogram. This will visualize the frequency of counties by their rent levels.

```{r}
# histogram of rent
ggplot(data = data, aes(x = MedianConRent)) +  # defines the plot space
  geom_histogram(       # defines what kind of visualization 
    binwidth = 50,      # each bar represents this increment
    fill = "red",       # bar color
    color = "gray50") + # lorder color
  labs(          
    title = "Median Contract Rent by County, 2023", # add text
    caption = "Source: ACS 5 Year Estimates",
    x = "Rent ($)",
    y  = "Counties"
  ) + 
  theme_minimal()  # choose a theme
```

The shape of the histogram provides useful storytelling information. We can see that the data have a generally normal (or 'bell curve') distribution with most counties hovering around the national average (of \$726). But there are also some very extreme outliers on the right side of the histogram. Those outliers would be the places with the most pressing housing affordability problem.

```{r}
# histogram of rent
ggplot(data = data, aes(x = MedianConRent)) +  
  geom_histogram(binwidth = 50, fill = "red",color = "gray50") + 
  geom_vline(aes(xintercept = 1742),   ## adding a line to show Brooklyn rent
             color = "blue",
             linetype = "dashed",
             size = .75) +
  geom_text(aes(x = 1950,     ## and labeling the line
                label = "Brooklyn", 
                y = 300), 
            color = "blue", 
            angle = 0, 
            vjust = 1.5) + 
  labs(          
    title = "Median Contract Rent by County, 2023", # add text
    caption = "Source: ACS 5 Year Estimates",
    x = "Rent ($)",
    y  = "Counties"
  ) + 
  theme_minimal()  # choose a theme
```

#### Choropleth maps

So where are these outliers? To find them, we can map the geographic pattern of our variables. This will identify regions of high/low values in each of the variables. To do this, we'll create a set of [choropleth maps](https://www.axismaps.com/guide/choropleth). This is a common data viz type that uses the color of map objects to visualize their data values.

```{r}
# first we'll drop Alaska, Hawaii, and overseas territories to map only the lower 48 states
## FIPS codes to exclude
exclude_list <- c("02", "15", "60", "66", "69", "72", "78")

# filter the data
lower48 <- data %>%
  filter(!str_sub(GEOID, 1, 2) %in% exclude_list)
```

```{r}
# now map median rent
ggplot(data = lower48) +  # defines the plot space
  geom_sf(aes(fill = MedianConRent), color = NA) +  # viz type = map
  coord_sf(crs = "ESRI:102010") +   # a relevant map projection for the region
  scale_fill_distiller(palette = "Reds", # define color fill
                       direction = 1,  # ramp from light to dark
                       name = "Rent ($)",  # label the legend
                       na.value = "gray90") + # color for nulls 
  labs(          
    title = "Median Contract Rent by County, 2023", # add text
    caption = "Source: ACS 5 Year Estimates"
  ) + 
  theme_minimal()  # choose a theme
  
```

As you can see, the outliers with very high rent are primarily urban counties in coastal states. The counties in New York are certainly among them. So we'll zoom into data focused one county: Brooklyn (aka Kings County).

```{r}
# set up an API query
data_NYC <- get_acs(geography = "tract",    # census tracts
                state = "New York",     # now we specify a geographic search
                county = "Kings",
                variables = variables,   # all variables defined in the list above
                output = "wide",         # row = observation, column = variable
                year = 2023,             # most recent year
                geometry = TRUE         # include spatial boundary data   
                ) 
```

Now we have a new data frame in the environment, called `data_NYC`. It has the same variables as the national data, but formatted instead for census tracts in Brooklyn.

```{r}
# do the same cleaning as before
data_NYC <- data_NYC %>%
  rename_with(~ gsub("E$", "", .x), .cols = everything()) %>%
  mutate(
    NewHomes_pct =  (HouseUnits_built2020orlater / HouseUnits_sum) * 100,
    NewComers_pct = (Owner_MovedIn_after2021+ Renter_MovedIn_after2021) /
      OccupiedHouseUnits_sum * 100)
```

```{r}
# and calculate descriptive statistics just to verify it worked
mean(na.omit(data_NYC$MedianConRent))
mean(na.omit(data_NYC$NewHomes_pct))
mean(na.omit(data_NYC$NewComers_pct))
```

```{r}
# median rent for one borough
ggplot(data = data_NYC) +
  geom_sf(aes(fill = MedianConRent), color = NA) +  # Changed the variable
  coord_sf(crs = "EPSG:32618") +  
  scale_fill_distiller(palette = "Reds",
                       direction = 1,
                       name = "Rent ($)",
                       na.value = "gray90") +  
  labs(
    title = "Median Contract Rent",
    subtitle = "By census tract in Brooklyn, 2020-2023",
    caption = "Source: ACS 5 Year Estimates"  
  ) +
  theme_minimal()
```

#### Scatterplots

So what might explain the housing affordability challenge in Brooklyn? Is it linked to supply factors like the availability of new homes, or demand factors like the prevalence of recent moves? We'll create some scatterplot, a visualization technique that plots one variable against the other, so you can see the overall pattern.

```{r}
# create a scatterplot for new housing
ggplot(data = data_NYC,   # define the data space
       aes(x = NewHomes_pct,  # x variable
           y = MedianConRent,  # y variable
           size = HouseUnits_sum)) +   # size of bubble, just for contet
  geom_point(color = "green", # color as name or RBG vector
             alpha = .5) +  # transparency
  scale_size_continuous(labels = scales::comma) +
  labs(
    title = "Does new construction relate to median rent?",
    subtitle = "In Brooklyn, 2020-2023",
    y = "Median Contract Rent ($)", 
    x = "New Homes (%)", 
    size = "Total Housing Units"
  )
```

```{r}
# create a scatterplot for recent movers
ggplot(data = data_NYC,   # define the data space
       aes(x = NewComers_pct,  # x variable
           y = MedianConRent, # y variable
           size = OccupiedHouseUnits_sum)) +  
  geom_point(color = "purple", # color as name or RBG vector
             alpha = .5) +  # transparency
  scale_size_continuous(labels = scales::comma) +
  labs(
    title = "Does migration relate to median rent?", 
    subtitle = "In Brooklyn, 2020-2023",
    y = "Median Contract Rent ($)", 
    x = "Recently Moved In (%)", 
    size = "Occupied Housing Units"
  )
```

On these plots, each dot represents a county. The dots' position represent their values of the x and y variables. The size of each dot represents the total 'population' in each county (in this case, based on housing units rather than people). This size parameter is included simply for context, since populous cities presumably have different kinds of housing dynamics than rural areas.

The point cloud tells us very useful information about the relationship between the two variables. Specifically, we can tell stories about:

1.  The relative **trend in the cloud**. Trending upward (from bottom left to top right) indicates a positive relationship. Trending downward indicates a negative relationship.

2.  The relative **dispersion of the cloud**. A tightly clustered cloud indicates a strong association between variables. A randomly distributed cloud indicates a weak association.

# Part 3: Statistically test data patterns

While visualization is useful, it only *describes* a data pattern. For data storytelling to be persuasive, however, we need to present evidence that tests the significance of the pattern and evaluates uncertainty around the analysis.

#### Correlation

Correlation tests are used to assess whether two variables are associated with each other. Thinking back to the scatterplot, these tests are assessing the direction of the point cloud (whether it's positive or negative) and the relatively compactness of the clould (more compact clouds indicate more closely related variables). The null hypothesis is that the variables are not associated. The alternative hypothesis is that they are.

For data with our variables' structure, we'll use a test called [Pearson's r correlation coefficient](https://statsthinking21.github.io/statsthinking21-core-site/ci-effect-size-power.html#pearsons-r). It will generate a correlation coefficient that ranges from -1 to 1.

-   0 indicates no correlation.

-   The directionality of the coefficient indicates whether the correlation is positive or negative (the direction the point cloud slopes toward).

-   The relative magnitude of the coefficient indicates the strength of correlation (the relative compactness or dispersion of the point cloud).

```{r}
# correlation between rent and new construction 
cor.test(x = data_NYC$NewHomes_pct,   # x variable          
         y = data_NYC$MedianConRent,  # y variables          
         method = "pearson",      # correlation method          
         use = "complete.obs")    # exclude rows with null values
```

Interpreting this, rent and home construction are positively but only weakly correlated, and that correlation is statistically significant at the 99% level.

```{r}
# correlation between rent and recent movers 
cor.test(x = data_NYC$NewComers_pct,           
         y = data_NYC$MedianConRent,          
         method = "pearson",          
         use = "complete.obs")
```

Interpreting this, rent and recent movers are positively correlated, and that correlation is statistically significant at the 99.9% level. The correlation is stronger for recent movers than for recent construction.

#### Linear models

A correlation test will tell us that two variables are associated. But, famously, *correlation does not equal causation*. For some hilarious illustrations of this point, see Tyler Vigen's [*Spurious Correlations*](https://www.tylervigen.com/spurious-correlations) blog.

To get closer to measuring causal patterns, we can use linear regression models to ask if one variable *predicts* the behavior of the other variable. The basic version of this is 'simple' [linear regression](https://statsthinking21.github.io/statsthinking21-core-site/fitting-models.html), which fits a line through the point cloud of your scatterplot in the most efficient manner possible.

The model is an algebraic line:

y = 𝛂 + 𝛃(x)

where:

-   y is the predicted value of y,

-   𝛂 is the intercept of the line. It can be interpreted as the value we would expect to see for y when the value of x is 0.

-   𝛃 is the slope of the line. It can be interpreted as how much we predict y to change if x increases by one unit.

```{r}
# define a regression model for new construction
lm(formula = MedianConRent ~ NewHomes_pct, data = data_NYC)
```

Interpreting this: we would expect a baseline rent of \$1732 (the intercept). For each 1% increase in new housing, we expect rent to increase by \$40 (the slope).

```{r}
# define a regression model for recent movers
lm(formula = MedianConRent ~ NewComers_pct, data = data_NYC)
```

Interpreting this: we start with a baseline rent of \$1365 (the intercept). For each 1% increase in recent movers, we expect rent to increase by \$42 (the slope).

#### Visualize the model

Finally, we'll add these model lines to the scatterplots we created before to visualize the overall trend. The `geom_smooth` layer will add a line using the same `y ~ x` formula we used above. It will also add a visualization of the confidence interval, which you can specify using a level parameter.

```{r}
# add model for new housing
ggplot(data = data_NYC,   
       aes(x = NewHomes_pct, y = MedianConRent)) +  
  geom_point(color = "green", alpha = .5) +  
  geom_smooth(method = "lm",  # use y ~ x
              level = 0.95,   # confidence interval around the line
              color = "red"
              ) +
  labs(
    title = "Does new construction relate to median rent?", 
    y = "Median Contract Rent ($)", 
    x = "New Homes as percent of market"
  )
```

```{r}
# create a scatterplot for recent movers
ggplot(data = data_NYC, aes(x = NewComers_pct, y = MedianConRent)) +  
  geom_point(color = "purple", alpha = .5) +  
  geom_smooth(method = "lm",  # use y ~ x
              level = 0.95,   # confidence interval around the line
              color = "orange"
              ) +
  labs(
    title = "Does in-migration relate to median rent?", 
    y = "Median Contract Rent ($)", 
    x = "Recently Moved In (%)"
  )
```

# Conclusion

To sum up, our goals for the workshop were to:

1.  Query American Community Survey data from the Census API. We did this using tools from the `tidycensus` library including `load_variables()` to view Census metadata and `get_acs()` to build and submit an API query.
2.  Visualize patterns in the data using histograms, choropleth maps, and scatterplots. We did this using the `ggplot2` library, which applies a 'grammar of graphics' approach to construct data visualizations. The key elements of our visualizations happened in how we used the `aes()` parameter to define x and y variables for the plot space (for the maps x = longitude and y = latitude, for the scatterplots, x = our predictor variables and y = median rent).
3.  Analyze whether those patterns are statistically significant using correlation and simple linear regression. We tested both using 'base R' functions – functions available as part of the core R language, not as an additional library. We used `cor.test()` to test whether the x and y variables were related to each other. Then we used `lm()` to define a simple linear regression model to test whether the x variable predicts behavior in the y variable.
